---
title: "Iteration and Annotation"
subtitle: "Prompt Engineering"
author:
  - name: Mick McQuaid
    email: mcq@utexas.edu
    affiliations: University of Texas at Austin
date: last-modified
date-format: "DD MMM YYYY"
format: 
  clean-revealjs:
    self-contained: true
    slide-number: false
    logo: quarto.png
    footer-logo-link: "https://quarto.org"
    sc-sb-title: false
    header-logo: iSchoolLogoLight.png
    header-logo-link: "https://ischool.utexas.edu"
    subtitle-as-header: false
    controls: true
    controls-layout: edges
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
filters:
  - reveal-header
  - autofade.lua
slide-level: 2
number-sections: false
engine: knitr
embed-resources: true
bibliography: master.bib
monofont: JetBrainsMono Nerd Font
mathfont: TeX Gyre Math
menu:
  side: 'right'
css: "style.css"
jupyter: python3
execute: 
  echo: true
---

# [Week FOUR]{.r-fit-text}

# Agenda
- Presentation: Shreya K
- News
- Review whatiknow
- Q&A on eB
- Q&A on m1
- Simple chatbot
- Iteration
- Annotation
- Work time

# Presentations

# News

# WhatIKnow

## Issues with the doc
- Several people dated their contributions! Let's all do that
- Nothing in the Intro tab? (shouldn't it introduce prompt engineering?)
- How about more of your opinions / reflections on what you read?
- The FAQ is empty! Don't you have questions?
- Some items need a TLDR 
- Are you finding it useful?

# eB

## Specification
- Use gen AI to assemble a bibliography in BibTeX format
- Manually assemble a bibliography at the ACM Digital Library
- Compare them
- The subject is PDF remediation: making a PDF accessible to people with disabilities
- I'm familiar with this literature, making it easier for me to detect hallucinations
- I can also judge the importance of the papers listed
- A recent bibliography in this area listed 39 sources, of which only about 20 were relevant
- Document your process!

# m1

## Deliverable
- A short qmd / html document describing the domain
- Use the template!
- The doc should include a discussion of the possible datasets / skills / mcp servers you might use (the actual dataset will be due in m2)
- You are not required to stick with the directions you give here, but this should be your current best guess of what you plan to do
- Examples: chatbot to emulate a foreign leader; chatbot to triage banking problems; chatbot to analyze tweets

# Simple chatbot

## First, a Hello World in R 

```{r}
# install.packages("pacman")
pacman::p_load(ellmer)
client <- chat_google_gemini()
client$chat("Summarize the plot of Romeo and Juliet in 20 words or fewer.")
```

Now `client` is an object (you can name it anything) and `chat` is a component of that object.

## Continuing the same conversation

```{r}
client$chat("Now do Hamlet.")
```

... and it remembers the question from the previous call.

## Same thing in Python
First, say `pip install chatlas` in a terminal.

```{python}
from chatlas import ChatGoogle
client = ChatGoogle(model = "gemini-2.5-flash")
response = client.chat("Summarize the plot of Romeo and Juliet in 20 words or fewer.",echo="none")
print(response)
```

Notice that the same exact answer may not be provided. This is a completely different conversation from the one in R.

## This Python conversation can also be continued.

```{python}
response = client.chat("Now do Hamlet.",echo="none")
print(response)
```

# Now a simple chatbot

```{r}
client =
  chat_google_gemini(
  system_prompt =
    ellmer::interpolate_file("posit-expense-policy.md"))
client$chat("I am a Posit employee. Based on the Posit Expense Policy, can I expense a hotel for business?")
```
## Cost
Anthropic isn't too good about itemizing costs, so I'm going to guess that the above, after running many times, cost about 0.25 USD. If you want to know the cost per conversation, you have to keep looking at Anthropic's up to the minute meter.

## Your turn
- Do the same thing in Python, using whatever key you have (OpenAI, Gemini, Anthropic, etc.)
- Put the key in an `~/.Renviron` file
- First task: Find something (not an expense report, but you can use my `posit-expense-policy.md` to get started) to put in the system prompt
- Can be in any domain you prefer, e.g., your project focus
- Remember to use `.` instead of `$` and ChatAnthropic instead of chat_anthropic (or whatever---I'll show the [possible models](https://posit-dev.github.io/chatlas/get-started/models.html))

# Iteration and annotation
For iteration, I think we can just brainstorm a lot of prompts and use a `foreach` to run them through `chatlas` in different conversations. Remember that each time we initialized the variable `client`, we were starting a new conversation where we could reuse that variable. We could instead give the conversations different names and run them in parallel, skipping back and forth as we get new ideas.

As for annotation, the effectiveness of system prompts makes it seem less pressing. Last semester, I asked students to annotate data for their projects. This semester, I'm not going to ask the same.

# Iteration

## How do you iterate?
- Do you document what you do?
- Do you use any prompt engineering tool to redeploy queries?
- How many times do you iterate?
- How do you know when you're done?
- There are many such worthwhile questions!
- Let's play with a prompt engineering tool to learn more

## Agenta
- [https://github.com/Agenta-AI/agenta](https://github.com/Agenta-AI/agenta)
- Let's pause to see a couple of Agenta videos
- $\langle$ agenta intro $\rangle$
- $\langle$ agenta quick start $\rangle$
- Let's try the workflow in the second video
- You can install it on your machine but I'll use the cloud version
- How many have Docker already installed?

## Example workflow
- Visit [https://github.com/Agenta-AI/agenta](https://github.com/Agenta-AI/agenta)
- Select Agenta Cloud
- Set up an account
- I tried using `example_tweets_test` without success, so I switched to `example_tweets_test_tiny`
- Follow me through the workflow

# Annotation

## Example
- We'll work with *Potato*, a current experimental annotation tool
- It's described in a recent paper, @Pei2023
- It's an open source project
- It's written in Python
- It's available on GitHub
- paper at [https://arxiv.org/abs/2212.08620](https://arxiv.org/abs/2212.08620)

## Getting started
- Clone the repository
- Install the dependencies
- Run the tool
- Sounds simple, right?
- Unfortunately, the instructions don't work
- I think they are out of date
- We'll do it our own way

## Part one

```bash
mkdir labeling && cd labeling
git clone https://github.com/davidjurgens/potato.git
cd potato
pip install -r requirements.txt
python potato/flask_server.py start project-hub/politeness_rating/configs/politeness.yaml -p 8000
```

## Part two
- Visit http://localhost:8000
- You should see a login screen
- You have to create an account
- Click on the "Create an account" link
- Enter an account name and a password, it doesn't really matter what

## Part three
- Log in with your new account
- You should see an intro, then an instance to annotate
- There should be fifty instances to annotate
- Try to annotate them all
- You can just enter a number from 1 to 5, it doesn't matter what
- Try not to give them all the same number

## Part four
- Do the demographics; again, they don't matter, just enter something
- Click on the "Submit" button
- Now you want to see your results
- They are in a folder called `potato/project-hub/politeness_rating/annotation_output/full/<your account name>/`
- Kill the server with `Ctrl-C` and use that window to navigate to the folder

## Part five
- There are several files in the folder
- `annotation_order.txt` is a list of the instances you annotated in order
- `assigned_user_data.json` is a list of the instances you annotated, with the annotations you made
- `annotated_instances.jsonl` is a list of the instances you annotated, with the annotations you made, in JSON Lines format
- Look at each file in a text editor

## Part six
- Convert the JSON Lines file to a CSV file
- Use the CSV file to calculate the average rating
- You may have some trouble extracting the scale
- I used `vi` to change the word `scale_1` to `scale`
- To do this to all five scale labels, I used the `vi` command `:%s/scale_[1-5]/scale/`
- Then I used the `extractIDandScale.py` script to extract the ID and scale from the JSON Lines file
- You may use an LLM for these tasks

## Part seven
- I used `vi` to get rid of lines that start with *Politeness*
- I used the `vi` command `:g/^Politeness/d`
- Next I used the `calcAvg.py` script to calculate the average rating
- However, average ratings are not very useful for Likert scales
- Instead, I made a stem-and-leaf plot
- It took several tries for gpt-4o to make it like the one I can easily generate in R

## Stem-and-leaf plot

```
  5 | 000000
  4 | 00000000000
  3 | 00000000000000000
  2 | 00000000000
  1 | 00000
```

See if you can make a stem-and-leaf plot for your data with an LLM

# [END]{.r-fit-text}

# References

::: {#refs}
:::

# Colophon

This slideshow was produced using `quarto`

Fonts are *Roboto Light*, *Roboto Bold*, and *JetBrains Mono Nerd Font*

