---
title: "exercise D"
subtitle: "Statistical Analysis"
date: today
author: Mick McQuaid
format:
  html:
    toc: true
    embed-resources: true
mainfont: TeX Gyre Schola
monofont: JetBrainsMono Nerd Font
mathfont: TeX Gyre Schola Math
---

# Intro

This documents my attempts to use an LLM to assist in analyzing a well-known dataset.

## Instructions

### Step one
Describe the Ames Housing dataset. This should include statistics and graphics describing the individual variables and the relationships between pairs of variables.

*You should use R to do this. You should ask the LLM to provide R scripts to do this, then include the R scripts in your .qmd file. Keep in mind that you should enclose the R code in triple backticks, like this: ` ```{r} ... ``` `*

### Step two
Make inferences about the determinants of the sale price of a house. This should include a regression analysis of the variables that appear to be most important.

*You should use the `lm()` function in R to form a regression model.*

### Step three
You should create four diagnostic plots of the final regression model you select. You should include the plots and a brief analysis of the plots.

*You should use the `plot()` function in R to form the plots. If you have created a model using a function call like `m <- lm(...)` then you can use `plot(m)` to create the four plots.*

### Step four
You should include your final judgment as to which is the best model, which constitutes a list of variables that determine the sale price.

### Conclusion
You should reflect on your work and the utility of the LLM in this task.

*Be sure to remove the above instructions before you turn in the file!*

# Step one: Description of data

## Load and preview the data

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(corrplot)
library(tidyr)

# Load the Ames Housing dataset
ames <- read.csv("../../resources/claude-code/amesHousing2011.csv")

# Display basic information about the dataset
cat("Dataset dimensions:", dim(ames)[1], "rows and", dim(ames)[2], "columns\n")
cat("\nFirst few rows:\n")
head(ames)

# Display structure of the dataset
str(ames)
```

## Summary statistics for key variables

```{r}
# Summary statistics for the target variable (SalePrice)
summary(ames$SalePrice)

# Summary statistics for key numeric variables
key_numeric <- c("LotArea", "YearBuilt", "OverallQual", "OverallCond",
                 "GrLivArea", "TotalBsmtSF", "FullBath", "BedroomAbvGr",
                 "GarageCars", "SalePrice")

summary(ames[, key_numeric])
```

## Distribution of Sale Price

```{r}
# Histogram of Sale Price
ggplot(ames, aes(x = SalePrice)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Sale Prices",
       x = "Sale Price ($)",
       y = "Frequency") +
  theme_minimal() +
  scale_x_continuous(labels = scales::comma)

# Log-transformed Sale Price
ggplot(ames, aes(x = log(SalePrice))) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "black") +
  labs(title = "Distribution of Log(Sale Price)",
       x = "Log(Sale Price)",
       y = "Frequency") +
  theme_minimal()
```

## Key categorical variables

```{r}
# Distribution of Overall Quality
ggplot(ames, aes(x = factor(OverallQual))) +
  geom_bar(fill = "coral") +
  labs(title = "Distribution of Overall Quality Ratings",
       x = "Overall Quality (1-10)",
       y = "Count") +
  theme_minimal()

# Distribution of Neighborhood
neighborhood_counts <- ames %>%
  count(Neighborhood) %>%
  arrange(desc(n))

ggplot(neighborhood_counts, aes(x = reorder(Neighborhood, n), y = n)) +
  geom_bar(stat = "identity", fill = "purple") +
  coord_flip() +
  labs(title = "Number of Houses by Neighborhood",
       x = "Neighborhood",
       y = "Count") +
  theme_minimal()
```

## Continuous variables exploration

```{r}
# Living Area vs Sale Price
ggplot(ames, aes(x = GrLivArea, y = SalePrice)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Sale Price vs. Living Area",
       x = "Above Ground Living Area (sq ft)",
       y = "Sale Price ($)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

# Year Built vs Sale Price
ggplot(ames, aes(x = YearBuilt, y = SalePrice)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Sale Price vs. Year Built",
       x = "Year Built",
       y = "Sale Price ($)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

# Overall Quality vs Sale Price
ggplot(ames, aes(x = factor(OverallQual), y = SalePrice)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Sale Price by Overall Quality",
       x = "Overall Quality (1-10)",
       y = "Sale Price ($)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

## Correlation analysis

```{r}
# Select numeric variables for correlation analysis
numeric_vars <- ames %>%
  select_if(is.numeric) %>%
  select(-Order, -PID) %>%  # Remove ID columns
  na.omit()

# Calculate correlation matrix
cor_matrix <- cor(numeric_vars)

# Find variables most correlated with SalePrice
saleprice_cor <- cor_matrix[,"SalePrice"]
saleprice_cor_sorted <- sort(abs(saleprice_cor), decreasing = TRUE)
cat("Top 10 variables correlated with SalePrice:\n")
print(head(saleprice_cor_sorted, 10))

# Visualize correlation matrix for top variables
top_vars <- names(head(saleprice_cor_sorted, 12))
cor_subset <- cor(numeric_vars[, top_vars])

corrplot(cor_subset, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Top Variables",
         mar = c(0,0,1,0))
```

## Relationships between key variable pairs

```{r}
# Total Basement SF vs Sale Price
ggplot(ames, aes(x = TotalBsmtSF, y = SalePrice)) +
  geom_point(alpha = 0.5, color = "brown") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Sale Price vs. Total Basement Area",
       x = "Total Basement Area (sq ft)",
       y = "Sale Price ($)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

# Garage Cars vs Sale Price
ggplot(ames, aes(x = factor(GarageCars), y = SalePrice)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Sale Price by Garage Capacity",
       x = "Number of Cars in Garage",
       y = "Sale Price ($)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

# Living Area vs Overall Quality
ggplot(ames, aes(x = GrLivArea, y = SalePrice, color = factor(OverallQual))) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_d() +
  labs(title = "Sale Price vs. Living Area by Overall Quality",
       x = "Above Ground Living Area (sq ft)",
       y = "Sale Price ($)",
       color = "Overall\nQuality") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

# Step two: Inferences about the determinants of the sale price

## Baseline model: Living Area only

```{r}
# Start with a simple model using just living area
model1 <- lm(SalePrice ~ GrLivArea, data = ames)
summary(model1)
```

The baseline model shows that living area alone explains a significant portion of sale price variance.

## Model 2: Adding Overall Quality

```{r}
# Add overall quality rating
model2 <- lm(SalePrice ~ GrLivArea + OverallQual, data = ames)
summary(model2)

# Compare models
anova(model1, model2)
```

Adding Overall Quality significantly improves the model, as evidenced by the increased R-squared and significant ANOVA test.

## Model 3: Adding Year Built and Basement Area

```{r}
# Add year built and total basement square footage
model3 <- lm(SalePrice ~ GrLivArea + OverallQual + YearBuilt + TotalBsmtSF,
             data = ames)
summary(model3)

# Compare with previous model
anova(model2, model3)
```

## Model 4: Adding Garage and Bathroom Features

```{r}
# Add garage capacity and bathrooms
model4 <- lm(SalePrice ~ GrLivArea + OverallQual + YearBuilt + TotalBsmtSF +
             GarageCars + FullBath + HalfBath, data = ames)
summary(model4)

# Compare with previous model
anova(model3, model4)
```

## Model 5: Adding Neighborhood (Categorical Variable)

```{r}
# Add neighborhood as a categorical predictor
model5 <- lm(SalePrice ~ GrLivArea + OverallQual + YearBuilt + TotalBsmtSF +
             GarageCars + FullBath + HalfBath + Neighborhood, data = ames)
summary(model5)

# Compare with previous model
anova(model4, model5)
```

## Comprehensive Model with Additional Features

```{r}
# Build a more comprehensive model
model6 <- lm(SalePrice ~ GrLivArea + OverallQual + OverallCond + YearBuilt +
             YearRemod.Add + TotalBsmtSF + X1stFlrSF + X2ndFlrSF +
             FullBath + HalfBath + BedroomAbvGr + KitchenAbvGr +
             TotRmsAbvGrd + Fireplaces + GarageCars + GarageArea +
             WoodDeckSF + OpenPorchSF + Neighborhood + BldgType,
             data = ames)
summary(model6)
```

## Checking for Multicollinearity

```{r}
# Calculate VIF (Variance Inflation Factor) to check for multicollinearity
library(car)
vif_values <- vif(model6)
print("VIF values for comprehensive model:")
print(vif_values)

# Identify problematic variables (VIF > 10)
high_vif <- vif_values[vif_values > 10]
if(length(high_vif) > 0) {
  cat("\nVariables with high multicollinearity (VIF > 10):\n")
  print(high_vif)
}
```

## Refined Model: Removing Multicollinearity Issues

```{r}
# Remove variables with high VIF (GarageArea and 1stFlrSF likely correlated with others)
model7 <- lm(SalePrice ~ GrLivArea + OverallQual + OverallCond + YearBuilt +
             YearRemod.Add + TotalBsmtSF + FullBath + HalfBath +
             BedroomAbvGr + TotRmsAbvGrd + Fireplaces + GarageCars +
             WoodDeckSF + OpenPorchSF + Neighborhood + BldgType,
             data = ames)
summary(model7)

# Check VIF again
vif_model7 <- vif(model7)
print("VIF values for refined model:")
print(vif_model7)
```

## Model Comparison Summary

```{r}
# Create a comparison table
library(broom)

model_comparison <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6", "Model 7"),
  Variables = c("GrLivArea",
                "GrLivArea + OverallQual",
                "Model2 + YearBuilt + TotalBsmtSF",
                "Model3 + GarageCars + Bathrooms",
                "Model4 + Neighborhood",
                "Comprehensive (20+ variables)",
                "Refined (removed multicollinearity)"),
  R_squared = c(
    summary(model1)$r.squared,
    summary(model2)$r.squared,
    summary(model3)$r.squared,
    summary(model4)$r.squared,
    summary(model5)$r.squared,
    summary(model6)$r.squared,
    summary(model7)$r.squared
  ),
  Adj_R_squared = c(
    summary(model1)$adj.r.squared,
    summary(model2)$adj.r.squared,
    summary(model3)$adj.r.squared,
    summary(model4)$adj.r.squared,
    summary(model5)$adj.r.squared,
    summary(model6)$adj.r.squared,
    summary(model7)$adj.r.squared
  ),
  AIC = c(
    AIC(model1),
    AIC(model2),
    AIC(model3),
    AIC(model4),
    AIC(model5),
    AIC(model6),
    AIC(model7)
  )
)

print(model_comparison)
```

## Interpretation of Final Model (Model 7)

```{r}
# Display coefficient summary for final model
coef_summary <- summary(model7)$coefficients
significant_coefs <- coef_summary[coef_summary[,4] < 0.05, ]

cat("Statistically significant predictors (p < 0.05):\n")
print(significant_coefs)

# Calculate confidence intervals for coefficients
cat("\n95% Confidence intervals for coefficients:\n")
print(confint(model7))
```

### Key Findings:

The regression analysis reveals that sale price is determined by multiple factors:

1. **Living Area (GrLivArea)**: Each additional square foot increases sale price
2. **Overall Quality (OverallQual)**: Higher quality ratings significantly increase price
3. **Year Built & Year Remodeled**: Newer homes command higher prices
4. **Basement Area (TotalBsmtSF)**: Additional basement space adds value
5. **Bathrooms**: Full and half bathrooms contribute to price
6. **Garage Capacity (GarageCars)**: More garage space increases value
7. **Neighborhood**: Location has a significant effect (categorical variable)
8. **Building Type**: Type of dwelling affects price

The final model (Model 7) achieves a high R-squared value, explaining a substantial portion of the variance in sale prices while avoiding multicollinearity issues.

# Step three: Regression diagnostics

## Diagnostic Plots for Final Model (Model 7)

The four standard diagnostic plots help us assess whether the regression assumptions are met:

```{r}
# Set up plotting area for 2x2 grid
par(mfrow = c(2, 2))

# Create the four diagnostic plots
plot(model7)

# Reset plotting area
par(mfrow = c(1, 1))
```

## Analysis of Diagnostic Plots

### 1. Residuals vs Fitted Plot (Top Left)

This plot checks for **linearity** and **homoscedasticity** (constant variance):

- **What to look for**: The residuals should be randomly scattered around the horizontal line at zero with no clear pattern
- **Observations**: The red smoothed line should be approximately horizontal and close to zero
- **Issues to watch for**:
  - Curved patterns suggest non-linearity
  - Funnel shapes indicate heteroscedasticity (non-constant variance)

```{r}
# Additional residual analysis
plot(model7$fitted.values, model7$residuals,
     main = "Residuals vs Fitted Values (Detailed)",
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch = 20,
     col = rgb(0, 0, 1, 0.3))
abline(h = 0, col = "red", lwd = 2)

# Add smoothed line
lines(lowess(model7$fitted.values, model7$residuals), col = "green", lwd = 2)
```

### 2. Normal Q-Q Plot (Top Right)

This plot checks if the **residuals are normally distributed**:

- **What to look for**: Points should fall approximately along the diagonal reference line
- **Observations**: Deviations at the tails indicate departure from normality
- **Issues to watch for**:
  - Points curving away from the line suggest skewed residuals
  - Heavy tails indicate outliers

```{r}
# Additional normality test
shapiro_test <- shapiro.test(sample(model7$residuals, 5000))
cat("Shapiro-Wilk normality test:\n")
cat("W =", shapiro_test$statistic, ", p-value =", shapiro_test$p.value, "\n")

# Histogram of residuals
hist(model7$residuals,
     breaks = 50,
     main = "Distribution of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     freq = FALSE)
curve(dnorm(x, mean = mean(model7$residuals), sd = sd(model7$residuals)),
      add = TRUE, col = "red", lwd = 2)
```

### 3. Scale-Location Plot (Bottom Left)

This plot checks for **homoscedasticity** (constant variance):

- **What to look for**: Points should be randomly scattered with a horizontal red line
- **Observations**: The spread of residuals should be roughly constant across fitted values
- **Issues to watch for**:
  - Upward or downward trends indicate heteroscedasticity
  - Fan shapes suggest variance changes with fitted values

```{r}
# Breusch-Pagan test for heteroscedasticity
library(lmtest)
bp_test <- bptest(model7)
cat("Breusch-Pagan test for heteroscedasticity:\n")
cat("BP =", bp_test$statistic, ", p-value =", bp_test$p.value, "\n")
if(bp_test$p.value < 0.05) {
  cat("Evidence of heteroscedasticity detected (p < 0.05)\n")
} else {
  cat("No significant evidence of heteroscedasticity (p >= 0.05)\n")
}
```

### 4. Residuals vs Leverage Plot (Bottom Right)

This plot identifies **influential observations** and **outliers**:

- **What to look for**: Points with high leverage and large residuals (outside Cook's distance contours)
- **Cook's Distance**: Values > 0.5 or > 1.0 suggest influential points
- **Issues to watch for**:
  - Points in upper-right or lower-right corners are influential
  - These points can disproportionately affect the regression coefficients

```{r}
# Identify influential observations
cooks_d <- cooks.distance(model7)
influential <- which(cooks_d > 4/length(cooks_d))

cat("Number of influential observations (Cook's D > 4/n):", length(influential), "\n")
if(length(influential) > 0) {
  cat("Influential observation indices:\n")
  print(head(influential, 10))

  # Show details of most influential observations
  top_influential <- order(cooks_d, decreasing = TRUE)[1:5]
  cat("\nTop 5 most influential observations:\n")
  print(ames[top_influential, c("SalePrice", "GrLivArea", "OverallQual",
                                 "YearBuilt", "Neighborhood")])
}

# Plot Cook's distance
plot(cooks_d, type = "h",
     main = "Cook's Distance",
     ylab = "Cook's Distance",
     xlab = "Observation Index")
abline(h = 4/length(cooks_d), col = "red", lty = 2)
```

## Additional Diagnostic Checks

### Checking for Non-linearity

```{r}
# Component + Residual plots for key continuous variables
library(car)

# Create component + residual plots for main predictors
crPlots(model7, terms = ~ GrLivArea + OverallQual + YearBuilt + TotalBsmtSF,
        main = "Component + Residual Plots")
```

### Summary of Regression Assumptions

```{r}
cat("=== Regression Diagnostics Summary ===\n\n")

# 1. Linearity
cat("1. LINEARITY: Checked via Residuals vs Fitted plot\n")
cat("   - Look for random scatter around zero\n\n")

# 2. Independence (cannot fully test without time series, but check)
library(car)
dw_test <- durbinWatsonTest(model7)
cat("2. INDEPENDENCE: Durbin-Watson test\n")
cat("   DW statistic:", dw_test$dw, "\n")
cat("   p-value:", dw_test$p, "\n")
if(dw_test$p > 0.05) {
  cat("   No significant autocorrelation detected\n\n")
} else {
  cat("   Warning: Possible autocorrelation in residuals\n\n")
}

# 3. Homoscedasticity
cat("3. HOMOSCEDASTICITY: See Breusch-Pagan test above\n\n")

# 4. Normality
cat("4. NORMALITY: See Shapiro-Wilk test and Q-Q plot above\n\n")

# 5. Multicollinearity
cat("5. MULTICOLLINEARITY: VIF values from Step Two\n")
mean_vif <- mean(vif_model7)
cat("   Mean VIF:", mean_vif, "\n")
if(mean_vif < 5) {
  cat("   Multicollinearity is not a concern (mean VIF < 5)\n\n")
} else {
  cat("   Warning: Possible multicollinearity issues (mean VIF >= 5)\n\n")
}

# Model performance metrics
cat("6. MODEL PERFORMANCE:\n")
cat("   R-squared:", summary(model7)$r.squared, "\n")
cat("   Adjusted R-squared:", summary(model7)$adj.r.squared, "\n")
cat("   Residual Standard Error:", summary(model7)$sigma, "\n")
cat("   F-statistic:", summary(model7)$fstatistic[1], "\n")
```

## Interpretation of Diagnostics

Based on the diagnostic plots and tests, we can assess the quality of our regression model:

1. **Linearity**: The Residuals vs Fitted plot shows whether the relationship between predictors and response is adequately captured by the linear model.

2. **Normality**: The Q-Q plot and Shapiro-Wilk test indicate whether residuals follow a normal distribution, which is important for inference (confidence intervals and hypothesis tests).

3. **Homoscedasticity**: The Scale-Location plot and Breusch-Pagan test reveal whether the variance of residuals is constant across all levels of fitted values.

4. **Influential Points**: The Residuals vs Leverage plot identifies observations that have unusual influence on the regression coefficients.

5. **Multicollinearity**: VIF values from Step Two ensure predictor variables are not too highly correlated with each other.

If any assumptions are violated, potential remedies include:
- **Non-linearity**: Transform variables (log, square root) or add polynomial terms
- **Heteroscedasticity**: Use weighted least squares or transform the response variable
- **Non-normal residuals**: Transform the response variable or use robust regression methods
- **Influential points**: Investigate these cases; consider removing outliers if justified
- **Multicollinearity**: Remove or combine correlated predictors

# Step four: results

$\langle$ replace this with your determination of the list of variables that determine sales price $\rangle$

# Conclusion

$\langle$ replace this with your reflections $\rangle$

# Addendum: Features of this file

*Note: delete this section before you turn in the file!*

-   Front matter
    -   Includes your name
    -   Includes the keyword "today" which resolves to the date on which you render the document
    -   Includes fonts---you should install these fonts on your computer or change the font specification to fonts you already have on your computer
    -   Includes the format (html) to which Quarto will render
    -   Includes some directives that are specific to that format: toc and embed-resources
    -   `toc` causes the table of contents to be rendered, on the right side of the frame by default
    -   `embed-resources` causes any diagrams to be included in the html file itself rather than linked---that way you can just submit the html file and I can view it instead of having to submit linked files
-   Headings: top level headings are preceded by a \# and a space; second level headings are preceded by \## and a space; you can go down several levels by increasing the number of \# symbols
-   Bulleted lists, formed by preceding the list with a blank line (or a heading) and beginning each line with a dash and a space (both are important)
-   LaTeX symbols, in this case $\langle$ and $\rangle$, which resolve to angle brackets when you render the document ... you can include any LaTeX math expressions between dollar signs or double dollar signs ... by the way, any dollar signs meant as real dollar signs should be preceded by a backslash, like \$ this, so Quarto doesn't get confused about whether you are starting an equation
-   Programmatic keywords, preceded and followed by a backtick, in this case, the name `eB.bib` bibliography file ... this causes the keyword to be rendered in a code font
-   Emphasis, by surrounding an *important* word with asterisks, causing it to be rendered in italics

Of course, you will delete all the instructions and comments in this file before you turn it in! I don't need to read them when I read your solution. The files you turn in (the qmd and the rendered html) will just include your work. These instructions and comments are just to help you get going.

